---
title: 'Bonanza juvenile hare counts and sizes: exploratory report'
author: "Kirsten Hodgson"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(effsize)
library(broom)
library(ggbeeswarm)
library(kableExtra)
library(here)
library(lubridate)
library(RColorBrewer)
```

```{r}
#read in data

hares_raw <- read.csv(here("data","bonanza_hares.csv"))
```

### Introduction

**This is where I will put my 4-5 sentence introduction to what this study is exploring, including basic background information about the subject, site, and study (with citations as needed), and a brief overview of what is included in the exploratory report.**

### Data and Analyses

**Here I will create a 3-4 sentence section where I will briefly describe the data, including citation and credit as needed, and provide an overview of the main analysis, statistical methods and tools used.**

### Exploratory Findings

#### Annual juvenile hare trap counts



```{r, fig.cap="The total number of trappings of juvenile hares at all grid sites. Trap counts are separated by study year between 1998 and 2012."}
#data wrangling

hares_juvenile_ann <- hares_raw %>% 
  mutate(date_new = mdy(date), .keep = "unused") %>% 
  mutate(year = year(date_new)) %>% 
  filter(age == "j")

#create table giving count of juvenile hares per year

hares_juv_ann_count <- hares_juvenile_ann %>% 
  group_by(year) %>% 
  summarize(annual_count = n())

#create a finalized data visualization of annual juvenile hare counts
ggplot(data = hares_juv_ann_count, aes(x = year, y = annual_count)) +
  geom_col(fill = "darkgoldenrod4") +
  labs(y = "Juvenile hare trap count", 
       title = "Annual juvenile hare trap counts at all grid sites, 1998-2012") +
  theme_light() +
  scale_x_discrete(name = "Year", 
                   limits = c(1998,1999,2000,2001,2002,2003,2004,2005,
                              2006,2007,2008,2009,2010,2011,2012))
```


Takeaways: The minimum count for juvenile hare trappings in the time period between 1998 and 2012 is zero. This occurred in three years: 1998, 2002, and 2009. The maximum count for juvenile hare trappings in the same time period is 126, in 1999. The median count is 13, and the mean is 25.3. There appears to be a general trend toward less juvenile hares trapped, with occasional peak years (1999, 2000, 2005, 2008). However, these data are directly influenced by effort, and more accurate analysis would depend on standardizing the trap counts based on number of days and number of traps per year. This could be accomplished by **ADD SOMETHING HERE**

#### Visualizing juvenile hare weights

```{r}
#data wrangling - creating a subset of juveniles grouped by grid and sex
hares_juv_weights <- hares_raw %>% 
  filter(age == "j") %>% 
  mutate(sex_new = case_when(sex == "f" ~ "Female",
                             sex == "m" ~ "Male",
                             TRUE ~ "Unknown")) %>% 
    group_by(grid, sex_new)

#summarize by grid and sex, excluding NA values
juv_weights_summary <- hares_raw %>% 
  filter(age == "j") %>% 
  filter(weight != "NA") %>% 
  mutate(sex_new = case_when(sex == "f" ~ "Female",
                             sex == "m" ~ "Male",
                             TRUE ~ "Unknown")) %>%
  group_by(grid,sex_new) %>% 
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight),
            sample_size = n())

#visualizing the data
p <- ggplot() +
  geom_beeswarm(data = hares_juv_weights, aes(x = sex_new, y = weight, color = sex_new),
                size = 1.5,
                alpha = 0.8) +
  scale_color_brewer(palette = "Dark2") +
  geom_point(data = juv_weights_summary, aes(x = sex_new, y = mean_weight)) +
  geom_errorbar(data = juv_weights_summary,
                aes(x = sex_new,
                    ymin = mean_weight - sd_weight,
                    ymax = mean_weight + sd_weight),
                width = 0.3) +
  labs(x = "Sex", y = "Weight (g)", 
       title = "Weight distribution of juvenile hares at grid sites", caption = "WRITE A CAPTION") +
  theme_light() +
  theme(legend.position = "none")

grid.labs <-c("Bonanza Black Spruce", "Bonanza Mature", "Bonanza Riparian")
names(grid.labs) <- c("bonbs", "bonmat", "bonrip")

p+facet_wrap(~grid, labeller = labeller(grid = grid.labs))
```

**Write 2-3 sentences describing the main takeaways here!!**

#### Comparing weights of male and female juvenile hares

```{r}
#data wrangling to get a subset exploring mean weights for juvenile male and female snowshoe hares
juv_m_f <- hares_raw %>% 
  filter(age == "j", sex %in% c("f", "m")) %>% 
  mutate(sex_new = case_when(sex == "f" ~ "Female",
                             sex == "m" ~ "Male",
                             TRUE ~ "Unknown")) %>% 
  group_by(sex_new)

```

```{r}
#creation of a summary table
juv_m_f_summary <- juv_m_f %>% 
  summarize(mean_weight = mean(weight, na.rm = T),
            sd_weight = sd(weight, na.rm = T),
            sample_size = n())
  
#Preparation of styled/finalized summary table
juv_m_f_summary %>% 
  kable(col.names = c("Sex", "Mean Weight (g)", "Standard Deviation (g)", "Sample Size"),
        caption = "Figure 3. Summary statistics for weight comparison of juvenile female and male   snowshoe hares in Bonanza Creek Experimental Forest, 1999-2012. Comparing the mean and standard deviation (g), and sample size of both groups.") %>% 
  kable_styling()
```

```{r, include = FALSE}
#Exploratory data viz for t-test assumptions
juv_m_f %>% 
  ggplot(aes(x = weight)) +
  geom_histogram(bins = 15) +
  facet_wrap(~sex_new)

juv_m_f %>% 
  ggplot(aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex_new)

#The female distribution looks pretty close to normally distributed, well within my level of comfort (sample size is also large, which helps). Male looks less closely to normally distributed, but still pretty good with the largest deviation at the smaller sizes. Because of the large sample size and central limit theorem, I feel comfortable to proceed with a two-sample t-test.
```

```{r}
#Preparation for and conducting a two-sample t-test and Cohen's *d*

female_juvenile_sample <- juv_m_f %>% 
  filter(sex == "f") %>% 
  pull("weight")

male_juvenile_sample <- juv_m_f %>% 
  filter(sex == "m") %>% 
  pull("weight")

t.test_m_f <- t.test(female_juvenile_sample, male_juvenile_sample)
cohen.d_m_f <- cohen.d(female_juvenile_sample, male_juvenile_sample, na.rm = TRUE)
```

A 2-3 sentence statement of the mean's comparison, including at least: the actual difference in means, Effect size (Cohen's *d*), and the outcome of a two-sample t-test - **USING IN-LINE VALUE REFERENCING**

#### Exploring the relationship between juvenile weight and hind foot length

```{r, include = FALSE}
#Data wrangling to create a subset of only juvenile hares
juveniles <- hares_raw %>% 
  filter(age == "j")

#Exploratory visualization
juveniles %>% 
  ggplot( aes(x = hindft, y = weight)) +
  geom_point()
```

#### Linear regression

The graph appears to show a somewhat linear relationship between weight and hind foot length, although the relationship is unclear at longer hind foot lengths. There are more outliers at longer lengths and it looks like there may be a trend toward higher variance (homoscedasticity may be a concern). 

```{r}
#set up for linear regression analysis - run linear model stored as hares_lm
hares_lm <- lm(weight ~ hindft, data = juveniles)
```

```{r}
#Get tidy versions of model output so that they can be called in text later
hares_lm_tidy <- broom::tidy(hares_lm)
hares_lm_glance <- glance(hares_lm)
```


```{r, include = FALSE}
# Look at model diagnostics to see if linear relationship is responsible
plot(hares_lm)
```

- Fitted vs. residuals - there is a trend toward higher variance, which makes me concerned about heteroscedasticity being a possibility
- QQ-plot - the residuals looks close to normally distributed except a few  points with a trend toward higher values and quite a few observations that are lower than the perfect line at lower values.
- Cook's Distance - don't even have a red dashed line for Cook's distance so no outliers to worry about

```{r}
#Get Pearson's r correlation
hares_cor <- cor.test(juveniles$weight, juveniles$hindft)

#Make a tidy version of the correlation output
hares_cor_tidy <- tidy(hares_cor)
```

```{r}
#Plot the data with a linear regression
ggplot(data = juveniles, aes(x = hindft, y = weight)) +
  geom_point(color = "darkgoldenrod4") +
  labs(x = "Hind foot length (mm)", y = "Weight (g)", 
       title = "Weight vs. hind foot length of juvenile snowshoe hares", 
       caption = "Figure 4. The relationship between juvenile snowshoe hare weight (g) and hind foot length (mm).") +
  geom_smooth(method = "lm", color = "gray30") +
  theme_light()
```

