---
title: 'Bonanza juvenile hare counts and sizes: exploratory report'
author: "Kirsten Hodgson"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(effsize)
library(broom)
library(ggbeeswarm)
library(kableExtra)
library(here)
library(lubridate)
```

```{r}
#read in data

hares_raw <- read.csv(here("data","bonanza_hares.csv"))
```

### Introduction

**This is where I will put my 4-5 sentence introduction to what this study is exploring, including basic background information about the subject, site, and study (with citations as needed), and a brief overview of what is included in the exploratory report.**

### Data and Analyses

**Here I will create a 3-4 sentence section where I will briefly describe the data, including citation and credit as needed, and provide an overview of the main analysis, statistical methods and tools used.**

### Exploratory Findings

#### Annual juvenile hare trap counts



```{r, fig.cap="The total number of trappings of juvenile hares at all grid sites. Trap counts are separated by study year between 1998 and 2012."}
#data wrangling

hares_juvenile_ann <- hares_raw %>% 
  mutate(date_new = mdy(date), .keep = "unused") %>% 
  mutate(year = year(date_new)) %>% 
  filter(age == "j")

#create table giving count of juvenile hares per year

hares_juv_ann_count <- hares_juvenile_ann %>% 
  group_by(year) %>% 
  summarize(annual_count = n())

#create a finalized data visualization of annual juvenile hare counts
ggplot(data = hares_juv_ann_count, aes(x = year, y = annual_count)) +
  geom_col(fill = "darkgoldenrod4") +
  labs(y = "Juvenile hare trap count", 
       title = "Annual juvenile hare trap counts at all grid sites, 1998-2012") +
  theme_minimal() +
  scale_x_discrete(name = "Year", 
                   limits = c(1998,1999,2000,2001,2002,2003,2004,2005,
                              2006,2007,2008,2009,2010,2011,2012))
```


Takeaways: The minimum count for juvenile hare trappings in the time period between 1998 and 2012 is zero. This occurred in three years: 1998, 2002, and 2009. The maximum count for juvenile hare trappings in the same time period is 126, in 1999. The median count is 13, and the mean is 25.3. There appears to be a general trend toward less juvenile hares trapped, with occasional peak years (1999, 2000, 2005, 2008). However, these data are directly influenced by effort, and more accurate analysis would depend on standardizing the trap counts based on number of days and number of traps per year. This could be accomplished by **ADD SOMETHING HERE**

#### Visualizing juvenile hare weights

```{r}
#data wrangling - creating a subset of juveniles grouped by grid and sex
hares_juv_weights <- hares_raw %>% 
  filter(age == "j") %>% 
  group_by(grid, sex)

#summarize by grid and sex, excluding NA values
juv_weights_summary <- hares_raw %>% 
  filter(age == "j") %>% 
  filter(weight != "NA") %>% 
  group_by(grid,sex) %>% 
  summarize(mean_weight = mean(weight),
            sd_weight = sd(weight),
            sample_size = n())

#visualizing the data
```

